\chapter{Code Generation and Runtime Support}
\label{ch:codegen-runtime}

After IR generation, the compiler emits ordinary C code for the selected field and target. This chapter describes the generated API, the byte-to-field mapping, the generated files, and the arithmetic implemented by the backend: data structures for field elements, load/store routines, arithmetic procedures, carry propagation, and loops corresponding to lowered reductions.

\section{Generated C API}
\label{sec:generated-c-api}

Every DSL hash function is emitted as a C function with the signature
\begin{lstlisting}[style=thesisC]
void h(uint8_t *output, uint8_t *key,
       uint8_t *message, size_t len);
\end{lstlisting}
The parameter \(len\) is a byte length at the C boundary. The generated function converts it to a number of field elements using the field chunk size \(c=\lfloor \pi/8\rfloor\). The DSL body then sees this derived field-element count.

\paragraph{Buffer convention.}
The caller is responsible for providing input buffers with enough padding for the generated loads. This convention keeps the inner loops simple: the compiler can compute the number of blocks as a ceiling division and may read slightly past the logical end of the message, as long as the caller has provided safe padding.

\paragraph{Output format.}
The output buffer is assumed to be large enough to hold the exported field element. The generated store routine writes the final reduced value in little-endian byte order. The number of exported bytes is \(\lceil \pi/8\rceil\), while the number of bytes consumed per input field element is \(\lfloor \pi/8\rfloor\). This asymmetry is intentional: input chunks avoid fractional bytes, while output must be able to represent a field element.

\section{Generated Files and Build Integration}
\label{sec:generated-files}

For a module named \(m\), the compiler emits a header \(m\texttt{.h}\), a source file \(m\texttt{.c}\), and shared data-structure files for the selected field representation. When automatic tests are enabled, it also emits a C++ test file. When performance generation is enabled, it emits a standalone C benchmarking source for the module.

The surrounding CMake project builds correctness and performance targets from the generated files. This build integration is deliberately thin. The compiler is responsible for producing ordinary C and C++ sources; CMake is responsible for compiling and linking them with handwritten tests and dependencies such as OpenSSL or Boost.

\section{Big Integer Representation}
\label{sec:bigint-representation}

Generated C code represents a field element as a struct containing one array entry per limb. Scalar field elements use machine-word limbs. Vectorized field elements use SIMD vector types as limb entries. Thus, the same high-level operation can be implemented by iterating over limbs, regardless of whether the limb type is scalar or vectorized.

The generated data structures also define constants derived from the field: the number of limbs, the limb mask, the size of the most significant limb, the input chunk size, and the number of bytes exported by a store. These constants are generated per field, so handwritten C tests and generated code use the same representation.

\Cref{lst:generated-structs} shows a generated data-structure fragment for the field \(2^{116}-3\) on a two-lane NEON target. In this configuration, \(\lambda=18\), \(\lambda'=8\), and \(\ell=7\). The scalar type stores one field element, while the vectorized type stores two field elements in limb-major form.

\begin{lstlisting}[style=thesisC,basicstyle=\scriptsize\ttfamily,caption={Generated scalar and vectorized field-element data structures.},label={lst:generated-structs}]
#include <arm_neon.h>

#define FIELD_PI 116ull
#define FIELD_THETA 3ull
#define FIELD_LAMBDA 18ull
#define FIELD_LAMBDA_PRIME 8ull
#define FIELD_LIMBS 7ull
#define FIELD_CHUNK_SIZE 14ull
#define FIELD_EXPORT_BYTES 15ull
#define FIELD_PRIME_HEX "0xffffffffffffffffffffffffffffd"

typedef struct {
    uint64_t limb0;
    uint64_t limb1;
    uint64_t limb2;
    uint64_t limb3;
    uint64_t limb4;
    uint64_t limb5;
    uint64_t limb6;
} bigint_t;

typedef struct {
    uint64x2_t limb0;
    uint64x2_t limb1;
    uint64x2_t limb2;
    uint64x2_t limb3;
    uint64x2_t limb4;
    uint64x2_t limb5;
    uint64x2_t limb6;
} vbigint_t;
\end{lstlisting}

\section{Load and Store}
\label{sec:load-store}

Loading is field-specific because it maps byte chunks into limbs. For scalar code, the load routine reads consecutive bytes, assembles them into an integer, and splits the result into limbs. For vectorized code, the generated load routine assembles several field elements at once, one per SIMD lane.

The important subtlety is that a loaded input field element contains only \(\lfloor \pi/8\rfloor\) bytes of data. If \(\pi\) is not divisible by eight, the remaining high bits of the field are not filled by consuming another byte. This convention matches the source-language view that buffer indices count field elements of a fixed byte chunk size.

Stores perform the reverse operation after a full reduction. The final field element is exported to bytes. The store operation is part of the IR for hash functions, so returning a field element from a hash does not require an extra wrapper call around generated code.

\section{Generated Arithmetic}
\label{sec:generated-arithmetic}

Addition is emitted as component-wise limb addition. In unsaturated representations, this is safe as long as the subsequent carry schedule keeps limb values within bounds.

\paragraph{Multiplication and squaring.}
Multiplication is emitted either as schoolbook multiplication or as Karatsuba multiplication. The schoolbook variant computes limb products and folds high terms back using the Crandall reduction identity. The Karatsuba variant reduces the number of limb multiplications at the cost of additional additions and higher register pressure. Which variant is faster depends on the field size and target.

Squaring has a dedicated implementation. It exploits symmetry of the product to avoid computing both \(a_i a_j\) and \(a_j a_i\) separately. The vectorized backends contain corresponding squaring code for NEON and AVX2.

For multiplication, the generated code follows the folded schoolbook formula from \Cref{sec:prime-field-arithmetic}. Each output limb accumulates ordinary products whose indices sum to that limb, plus wrapped products whose indices exceed the field size and are multiplied by the Crandall factor. The generated code is specialized enough that the loops are usually unrolled in the emitted C.

\section{Carry and Full Reduction}
\label{sec:carry-full-reduction}

Generated carry code propagates overflow bits from one limb to the next. The carry out of the most significant limb is multiplied by \(\theta\) and folded into the least significant limb. For vectorized code, the same operation is performed lane-wise using SIMD shifts, masks, and additions.

A full reduction is emitted at the end of hash functions. It handles edge cases where carry propagation alone leaves a value in the range above the prime. This final step is more expensive than an ordinary carry, but it is outside the main loop and therefore has less impact on throughput for long messages.

\section{Backends}
\label{sec:backends}

The scalar backend is the simplest backend and serves two roles. It is useful in its own right, and it provides a baseline for measuring vectorization speedups.

\paragraph{Vector backends.}
The NEON backend targets two 64-bit lanes. It is useful for ARM systems where two field elements can be processed in parallel. The AVX2 backend targets four 64-bit lanes. It follows the same limb-major layout, but the wider lane count changes the generated loads, lane-power generation, and vectorized loop structure.

\section{Performance Harness}
\label{sec:performance-harness}

The compiler can emit a small benchmark program for a generated hash function. The harness allocates key, message, and output buffers, calls the generated hash function for a range of byte lengths, and reports elapsed cycles per call. It treats the generated hash function as a black box: it depends only on the C API, not on the internal IR or on any static analysis.

The purpose of the harness is practical. It lets the same DSL module be measured under different fields, targets, multiplication algorithms, and carry strategies. The evaluation chapter will use this machinery to compare the generated implementations.

\section{Optional Formatting}
\label{sec:optional-formatting}

The compiler includes an optional formatting pass for generated C and C++ sources. Formatting is not part of the semantic pipeline, but it makes generated files easier to inspect, debug, and compare against handwritten tests. Since the generated code is intended to be fast but not necessarily readable, this convenience is useful despite being orthogonal to performance.
